---
title: "Introduction to boompy: R Interface to BOOM Python"
author: "Claude Assistant"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to boompy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The `boompy` package provides an R interface to the BOOM (Bayesian Object Oriented Modeling) Python library. BOOM offers a comprehensive suite of Bayesian statistical models and methods, now accessible through familiar R syntax.

## Installation

First, ensure you have Python (>= 3.7) installed with the BOOM Python package:

```{r install, eval=FALSE}
# Install the boompy R package
install.packages("boompy")

# Check installation
library(boompy)
boom_check_installation()
```

## Getting Started

### Linear Regression

The `boom_lm()` function provides Bayesian linear regression with a familiar R interface:

```{r linear-regression}
# Generate example data
set.seed(123)
n <- 100
x <- rnorm(n)
y <- 2 + 3*x + rnorm(n)
data <- data.frame(y = y, x = x)

# Fit linear regression
model <- boom_lm(y ~ x, data)
summary(model)

# Plot diagnostics
plot(model)
```

### Generalized Linear Models

The package supports various GLM families through `boom_glm()`:

```{r glm-examples}
# Logistic regression
binary_data <- data.frame(
  x = rnorm(200),
  y = rbinom(200, 1, prob = 0.5)
)
logit_model <- boom_glm(y ~ x, family = "binomial", data = binary_data)

# Poisson regression
count_data <- data.frame(
  x = rnorm(150),
  y = rpois(150, lambda = 2)
)
poisson_model <- boom_glm(y ~ x, family = "poisson", data = count_data)
```

### State Space Models

Bayesian structural time series (BSTS) models are available through `boom_bsts()`:

```{r state-space}
# Generate time series data
y <- cumsum(rnorm(100)) + rnorm(100)

# Fit local level model
ss_model <- boom_bsts(y, state.specification = list(type = "local_level"))
plot(ss_model)

# Forecast
forecast <- predict(ss_model, h = 10)
```

### Mixture Models

Finite mixture models can be fitted using `boom_mixture()`:

```{r mixture}
# Generate mixture data
y1 <- rnorm(100, mean = -2, sd = 1)
y2 <- rnorm(150, mean = 3, sd = 1.5)
y <- c(y1, y2)

# Fit Gaussian mixture
mix_model <- boom_mixture(y, k = 2)
summary(mix_model)
plot(mix_model)
```

### MCMC Sampling

The package provides flexible MCMC sampling capabilities:

```{r mcmc}
# Define log density for standard normal
log_density <- function(x) {
  -0.5 * sum(x^2)
}

# Run Metropolis-Hastings
samples <- boom_mcmc(log_density, initial = c(0, 0), niter = 2000, burn = 1000)
summary(samples)

# Plot results
plot(samples, type = "density")
```

## Model Comparison

Compare multiple models using information criteria:

```{r model-comparison}
# Fit competing models
model1 <- boom_lm(y ~ 1, data)         # Intercept only
model2 <- boom_lm(y ~ x, data)         # Linear
model3 <- boom_lm(y ~ x + I(x^2), data) # Quadratic

# Compare models
comparison <- boom_compare_models(model1, model2, model3)
print(comparison)
```

## Advanced Features

### Custom Priors

While the current implementation uses default priors, future versions will support custom prior specifications:

```{r custom-priors, eval=FALSE}
# Future feature: custom priors
prior <- list(
  beta = normal_prior(mean = 0, sd = 10),
  sigma = inverse_gamma_prior(shape = 1, rate = 1)
)
model <- boom_lm(y ~ x, data, prior = prior)
```

### Parallel Processing

For computationally intensive models, parallel processing can be enabled:

```{r parallel, eval=FALSE}
# Future feature: parallel MCMC chains
model <- boom_bsts(y, niter = 10000, parallel = TRUE, chains = 4)
```

## Performance Considerations

The BOOM Python backend leverages NumPy and SciPy for efficient numerical computations. For large datasets:

1. **Memory**: Data is passed between R and Python, so memory usage may be higher
2. **Speed**: Most computations occur in optimized Python/NumPy code
3. **Parallelization**: Future versions will support parallel MCMC chains

## Troubleshooting

If you encounter issues:

```{r troubleshooting, eval=FALSE}
# Check Python configuration
reticulate::py_config()

# Verify BOOM module
boom_version()

# Enable verbose output
options(boompy.verbose = TRUE)
```

## Conclusion

The `boompy` package brings the power of BOOM's Bayesian modeling to R users, combining:

- Familiar R syntax and workflows
- Powerful Python computational backend
- Comprehensive model diagnostics and visualization
- Seamless integration with R's data analysis ecosystem

For more examples and detailed documentation, see the package help files and additional vignettes.